The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-12-07 Paper.Rmd
finished editing Intro for this round, some items for Shya to take a look at, please

Well-developed statistical tools do exist, and are 
applied widely in other fields, that provide a demonstrably complete 
description of the dependence 
between two or more variables [@joe2014_dependence; @nelsen2006_copula]. 
The goal of this paper is to illustrate the potential for 
broader application of these
tools in ecology and related fields. The tools are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence between the variables). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}, panel A or B), the associated dependence
or copula structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}E-F, which show the copula structure of 
A-B, respectively. Ranking serves to uniformize the 
marginal distributions of the underlying random variables, isolating the 
depedence structure only.
Mathematically, if $(X,Y)$ is a continuous bivariate random
variable and the cummulative distribution functions (cdfs) of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence or copula structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in the unit square, 
$[0,1] \times [0,1]$, and has uniform marginal distributions. Densities
corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}E-F are in Fig.
\ref{fig_cop_pedag1}G-H, respectively. Strictly speaking, 
the copula is actually defined as the bivariate 
cdf of $(F_X(X),F_Y(Y))$ rather than as this random variable itself, 
but this makes little conceptual difference. See @nelsen2006_copula and @joe2014_dependence
for mathematical details. An important theorem of A. Sklar [@sklar1959_theorem] 
states that copulas give a unique and complete 
representation of dependence
between variables, in the sense that the joint distribution of $(X,Y)$ 
can be reconstructed completely from the 
copula and the marginal distributions. 
Copula theory is also well developed for
multivariate distributions.

@@

A principal benefit of a copula approach to ecological data is 
that such an approach can sensitively model and detect
tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of two variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence;
and \emph{vice versa} if larger values are more strongly associated than smaller
ones. Datasets of the same corelation can have a range of tail dependence 
characteristics (Fig. \ref{fig_cop_pedag}), which can be quantified (Methods).

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm,height=8 cm]{./Results/copula_pedagog_fig_BB1.pdf}
\caption{(A-C) Data generated using the BB1 copula (Methods) with three different 
sets of parameters chosen to produce a variety of 
lower-tail (LT) and upper-tail (UT) dependence strengths (Methods), but the same 
Kendall correlation (to two digits) in all cases. (D-F) Similar to A-C, but
with stronger correlation. Asymmetry of tail dependence (LT-UT) is the same 
within each column of panels.\label{fig_cop_pedag}}
\end{center}
\end{figure}
<!--***DAN: Shya, please reduce it to three panels in each row (the middle one and the most extreme ones), while keeping the panels the same size, and label the panels A-F. Please also get rid of the title over each panel, and express the UT and LT values to two digits, and make the font slightly bigger for those as well as for the axis labels and numbers. Please also remove the 
color - we don't need it and it would cost a lot of money.-->

One way we may expect a copula approach 
to better reflect ecological relationships has to do with Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, as well
as on other factors, data on soil nitrogen and resulting growth rates may be 
expected to look more like Fig. \ref{fig_cop_pedag1}A or E than like 
Fig. \ref{fig_cop_pedag1}B or F. Nitrogen 
will control plant growth, producing a strong
relationship, only when it is limiting, in the left portion of the N and 
plant-growth-rate distributions. 
Nitrogen will be less related or unrelated to growth rates in the right tails of 
the distributions, as pictured in Fig. \ref{fig_cop_pedag1}A and E. The portion
of the distributions that dominante the association is visible in the copula
approach, but may not be reflected by correlation and other standard techniques.

A second reason we may expect a copula-based approach to benefit 
ecological research is that such an approach may help illuminate 
causal mechanisms by which variables are related. For instance, if two species
X and Y are strong competitors, we may expect abundances of the two species
across quadrats or through time to be negatively related, as in 
Fig. \ref{fig_cop_pedag1}I-J. If X is plotted on the horizontal axis
and Y is plotted on the vertical, then the plot in Fig. \ref{fig_cop_pedag1}I 
suggests X is the dominant 
competitor: when X is abundant, Y is necessarily rare because it is suppressed 
by X; whereas when X is rare, Y
may be abundant or may also be relatively rare due to 
limiting factors other than X. 
Alternatively, Fig. \ref{fig_cop_pedag1}J suggests that Y
is the dominant competitor. The copula approach again reveals which portions of the
distributions are principally responsible for the association between variables.

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects (as seems 
frequently to be the case).
Spatial dependence in an environmental variable 
tends to beget dependence across space
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of spatial synchrony.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @walteretal2017].

A fourth reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of catches from 
local areas. We will explore how details, captured by copula methods, of the 
nature of the dependence between local quantities can influence the spatial 
mean or sum, and how this may influence human concerns and 
higher organizational levels in 
ecology. Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds in multiple wind sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively asscoiated and also strongly
related to each other in their right tails, then local outbreaks tend 
to occur together, creating regional epidemics. If populations
of a crop or other resource species are strongly 
related in their left tails, then 
episodes of local scarcity occur together, creating regional 
shortages. On the other hand left-tail dependence in the pest species 
and right-tail dependence in the crop, even if overall correlation were 
unchanged, would have much more benign effects.

<!--DAN: made it to here 2018 12 06
edit and add references to the following-->
Though a few examples have been found which shows copula approach to model dependencies in bioinformatics [@Kim2008], medical research [@Emura2016], a short period study of intersite dependence of rainfall data [@Serinaldi2008], a recent study on characterizing and assessing risks in drought management [@She2018] or monitoring changes in precipitation-extremes [@Goswami2018], the multivariate copula approach still, we believe, remains unexplored in ecology or related fields and thus has potential scope to better capture the evolution of ecological time series data.

 In this article, we analyzed the dependence structure for several bivariate and multivariate ecological, community and biogeochemical datasets. In addition to a copula based \nameref{Model_selection}, we also carried out a \nameref{NPA} with three different statistics to verify our model selection based results.

We want to address the following questions. \textbf{(Q1)} Do datasets in ecology and related fields have non-Normal copula srtructure? Do they show tail dependence distinct from that of a Normal copula, and in particular do they show asymmetric tail dependence? \textbf{(Q2)} If so, what are some possible causes of non-Normal copula structure? And \textbf{(Q3)} what are the consequences of non-Normal copula structure and tail dependence for our understanding of biology and for applications? 
In \nameref{Data}, we described briefly each of bivariate and multivariate datasets we used to answer these questions and we provide a detailed descriptions of the implemented methods in \nameref{Methods}. In the \nameref{Results} section, we summarized our findings for each dataset. The next two sections are focused on possible causes and consequences of non-Normal copula structure in ecology and related fields with applications shown in real data as well as some simulations. Finally, we conclude in the \nameref{Discussion} section.

<!--# Material and methods \label{Material&Methods}-->

>>>>

Well-developed statistical tools do exist, and are 
applied widely in other fields, that provide a demonstrably complete 
description of the dependence 
between two or more variables [@joe2014_dependence; @nelsen2006_copula]. 
The goal of this paper is to illustrate the potential for 
broader application of these
tools in ecology and related fields. The tools are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence between the variables). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the associated dependence
or copula structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}E-F, which show the copula structure of 
A-B, respectively. Ranking serves to uniformize the 
marginal distributions of the underlying random variables, isolating the 
depedence structure only.
Mathematically, if $(X,Y)$ is a continuous bivariate random
variable and the cummulative distribution functions (cdfs) of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence or copula structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in the unit square, 
$[0,1] \times [0,1]$, and has uniform marginal distributions. Densities
corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}E-F are in Fig.
\ref{fig_cop_pedag1}G-H, respectively. Strictly speaking, 
the copula is actually defined as the bivariate 
cdf of $(F_X(X),F_Y(Y))$ rather than as this random variable itself, 
but this makes little conceptual difference. See @nelsen2006_copula and @joe2014_dependence
for mathematical details. An important theorem of A. Sklar [@sklar1959_theorem] 
states that copulas give a unique and complete 
representation of the dependence
between variables, in the sense that the joint distribution of $(X,Y)$ 
can be reconstructed completely from the 
copula and the marginal distributions. 
Copula theory is also well developed for
multivariate distributions.

@@

A principal benefit of a copula approach to ecological data is 
that such an approach can sensitively model and detect
tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of two variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, E);
and \emph{vice versa} if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, F). Datasets of the same 
corelation can have a range of tail dependence 
characteristics (Fig. \ref{fig_cop_pedag}), which can be quantified (Methods).

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm,height=8 cm]{./Results/copula_pedagog_fig_BB1.pdf}
\caption{(A-C) Data generated using the BB1 copula (Methods) with three different 
sets of parameters chosen to produce a variety of 
lower-tail (LT) and upper-tail (UT) dependence strengths (Methods), but the same 
Kendall correlation (to two digits) in all cases. (D-F) Similar to A-C, but
with stronger correlation. Asymmetry of tail dependence (LT-UT) is the same 
within each column of panels.\label{fig_cop_pedag}}
\end{center}
\end{figure}
<!--***DAN: Shya, please reduce it to three panels in each row (the middle one and the most extreme ones), while keeping the panels the same size, and label the panels A-F. Please also get rid of the title over each panel, and express the UT and LT values to two digits, and make the font slightly bigger for those as well as for the axis labels and numbers. Please also remove the 
color - we don't need it and it would cost us a lot of money.-->

One way we may expect a copula approach 
to better reflect ecological relationships has to do with Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, as well
as on other factors, data on soil nitrogen and resulting growth rates may be 
expected to look more like Fig. \ref{fig_cop_pedag1}A or E than like 
Fig. \ref{fig_cop_pedag1}B or F. Nitrogen 
will control plant growth, producing a strong
relationship, only when it is limiting, in the left portion of the N and 
plant-growth-rate distributions. 
Nitrogen will be less related or unrelated to growth rates in the right tails of 
the distributions, as pictured in Fig. \ref{fig_cop_pedag1}A and E. The portion
of the distributions that dominate the overall association is visible in the copula
approach, but may not be reflected by correlation and other standard techniques.

A second reason we may expect a copula-based approach to benefit 
ecological research is that such an approach may help illuminate 
causal mechanisms by which variables are related. For instance, if two species
X and Y are strong competitors, we may expect abundances of the two species
across quadrats or through time within a single quadrat to be negatively related, as in 
Fig. \ref{fig_cop_pedag1}I-J. If X is plotted on the horizontal axis
and Y is plotted on the vertical, then the plot in Fig. \ref{fig_cop_pedag1}I 
suggests X is the dominant 
competitor: when X is abundant, Y is necessarily rare because it is suppressed 
by X; whereas when X is rare, Y
may be abundant or may also be relatively rare due to 
limiting factors other than X. 
Alternatively, Fig. \ref{fig_cop_pedag1}J suggests that Y
is the dominant competitor. The copula approach again reveals which portions of the
distributions are principally responsible for the association between variables, and this 
detail may help support or refute alternative mechanistic and causal hypotheses about 
the relationship between the variables.

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget dependence across space
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of spatial synchrony.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A fourth reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of catches from 
local areas. We will explore how details, captured by copula methods, of the 
nature of the dependence between local quantities can influence the spatial 
mean or sum, and how this may influence human concerns and 
higher organizational levels in 
ecology. Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks tend 
to occur together, creating regional epidemics. If populations
of a crop or other resource species are strongly 
related in their left tails, then 
episodes of local scarcity occur together, creating regional 
shortages. On the other hand left-tail dependence in the pest species 
and right-tail dependence in the crop, even if overall correlation were 
unchanged, would have much more benign effects.

We approached our overall goal of illustrating the potential of 
copula approaches for understanding dependences in ecology by addressing the 
folling specific questions. (Q1) Do datasets in ecology and related fields 
have copula structure distinct from a Gaussian/normal distribution (here called 
non-normal copula structure)? Do they show tail dependence distinct from that 
of a normal distribution, and in particular do they show asymmetric tail 
dependence? (Q2) If so, what are some possible causes/mechanisms of this 
non-normal copula structure? And (Q3) what are the consequences of non-normal 
copula structure and tail dependence for our understanding of biology and for 
applications? To address Q1, we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data. We selected these datasets to span multiple 
sub-fields and organizational levels of ecology, in an attempt to be 
reasonably representative of the field as a whole. The second section of the 
paper introduces the data, and the third and fourth sections use the data to 
address Q1. The fifth and sixth sections, which address Q2, use simple models to 
explore mechanisms. The seventh and eigth sections, which address Q3, use both
the data section two as well as models to explore a variety of consequences
of copula structure and tail dependence for applications and for our understanding
of ecology. The ninth section is the Discussion. 
Copula approaches have been used to great benefit in neuroscience [@onken2009], 
bioinformatics 
[@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], finance, and other fields. 
To our knowledge, these approaches have been used seldom or never for ecological 
research. We argue that the benefits to ecological research of a complete statistical 
desription of dependence will be substantial, and we hope this paper stimulates the
use of copula approaches across the discipline.
